The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Träff, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}

### 2014-01-26 article/article.tex
rework single choice balls into bins

A simple strategy for the balls-into-bins problem is to choose one bin uniformly and independently at random for each ball, which we will call the \emph{single-choice balls-into-bins} strategy. Using this strategy, the maximum load of the fullest bin will be $\frac{m}{n} + \Theta\left(\sqrt{\frac{m \log n}{n}}\right)$ w.h.p.~\cite{RS98}.

Consider a variation of the above \emph{single-choice balls-into-bins} strategy: For each ball, chose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}\footnote{If there are multiple bins with the same minimal load, chose one among them arbitrarily.}. This simple \emph{multiple-choice balls-into-bins} scheme, which we will call the \emph{Greedy} algorithm, achieves a maximum load of $\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d}\right)$ w.h.p.~\cite{ABKU99}~\cite{BCSV06}. This apperantly minor change introduces a small amount of choice which, compared to the single-choice scheme, leads to a significant reduction in the maximum load. 
\begin{comment}
Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. 
\end{comment}
Extensive research following the influental paper by Azar et. al \cite{ABKU99}, who were the first to explicitly prove this phenomenon, suggests that this \emph{two-choice paradigm} is not just an artificat of the simple balls-into-bins model, but a general principle applicable to a variety of problems.

@@

In the \emph{single-choice balls-into-bins} strategy, each ball is placed into a bin choosen uniformly and independently at random. \todo{more specific?}

The fullest bin contains $\Theta \left( \frac{\log n}{\log \log n} \right)$ for the lightly loaded case with high probability\footnote{We say that an event $\mathcal E$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E \right] = 1 - o(1)$, except when stated otherwise.} \cite{RS98}. Since the average load of the bins is 1, this is also a bound for the gap. 
%More accurately, the number of balls in the bin with the most balls is $\Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$ \cite{G91}. %\todo{really include Gonnet?}
For the \emph{heavily loaded case} the maximum load is $\frac{m}{n} + \Theta\left(\sqrt{\frac{m \log n}{n}}\right)$ w.h.p.

\todo{proof? Maybe just outline the $m = n$ case as in \cite{RS98}?}

>>>>

A simple strategy for the balls-into-bins problem is to choose one bin uniformly and independently at random for each ball, which we will call the \emph{single-choice balls-into-bins} strategy. Using this strategy, the maximum load of the fullest bin will be $\frac{m}{n} + \Theta\left(\sqrt{\frac{m \log n}{n}}\right)$ with high probability\footnote{We say that an event $\mathcal E$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E \right] = 1 - o(1)$, except when stated otherwise.}~\cite{RS98}.

Consider a variation of the above \emph{single-choice balls-into-bins} strategy: For each ball, chose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}\footnote{If there are multiple bins with the same minimal load, chose one among them arbitrarily.}. This simple \emph{multiple-choice balls-into-bins} scheme, which we will call the \emph{Greedy} algorithm, achieves a maximum load of $\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d}\right)$ w.h.p.~\cite{ABKU99}~\cite{BCSV06}. This apparently minor change introduces a small amount of choice which, compared to the single-choice scheme, leads to a significant reduction in the maximum load. 
\begin{comment}
Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. 
\end{comment}
Extensive research following the influental paper by Azar et. al \cite{ABKU99}, who were the first to explicitly prove this phenomenon, suggests that this \emph{two-choice paradigm} is not just an artificat of the simple balls-into-bins model, but a general principle applicable to a variety of problems.

@@

In the \emph{single-choice balls-into-bins} strategy, each ball is placed into a bin chosen uniformly and independently at random. In other words, a ball gets placed into a certain bin with probability $\frac{1}{n}$.

The fullest bin contains
\[
\Theta \left( \frac{\log n}{\log \log n} \right)
\] 
balls in the lightly loaded case w.h.p. Since the average load of the bins is 1, this is also a bound for the gap. 
%More accurately, the number of balls in the bin with the most balls is $\Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$ \cite{G91}. %\todo{really include Gonnet?}
In the \emph{heavily loaded case} the maximum load is 
\[
\frac{m}{n} + \Theta\left(\sqrt{\frac{m \log n}{n}}\right)
\] 
w.h.p.~\cite{RS98}.
